Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-10-15 paper/paper.Rmd
final corrections

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich Schiller University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    # must use German name, see https://www.lmu.de/de/die-lmu/struktur/zentrale-universitaetsverwaltung/kommunikation-und-presse/lmu-brand-guide/content-style-guide/index.html
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena \AND'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-Universität München'
    address: |
      | Department of Statistics
      | Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for machine learning in R with \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for machine learning in R with mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) as an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com), demonstrating its use in a step-by-step example.
  Currently various R packages implementing different spatiotemporal partitioning strategies exist: sperrorest, blockCV, CAST.
  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and common interface.
  This is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
  This simplifies the the integration of package-specific syntax into an overarching machine-learning pipeline, making it easier for users to choose from a variety of spatiotemporal resampling methods.
    The users' decision to use a specfic resampling method should depend on the predictive task at hand, the autocorrelation within the data, and the spatial structure of the sampling design or geographic objects being studied.
keywords:
  formatted: [cross-validation, predictive performance, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
  plain:     [cross-validation, predictive performance, autocorrelation, spatial, temporal, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{adjustbox}
  \usepackage{multirow}
  \usepackage{tabularx}
  \usepackage{placeins}
  \usepackage{scrextend}
  \usepackage{tablefootnote}
  \newcommand{\specialcell}[2][l]{%
    \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
tables: true
bibliography: paper.bib
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
---

@@

Although alternatives such as the bootstrap exist and have some advantages [@efron1983; @hand1997], cross-validation (CV) is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
In its basic form, it is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data, or underlying environmental processes inducing some sort of spatial coherence at the landscape scale.
In treating dependent observations as independent, or ignoring autocorrelation, CV test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias in assessing the transferability of flexible machine-learning models.

This CV bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
It is most easily understood from a predictive modeling perspective by focusing on the question where (and when) the model should be used for prediction.
In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain [*intra-domain* prediction, @brenning2005] --- not *within* the same field, which obviously presents only a single crop type that is already known from the training sample.
In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if we wanted to predict within a field.
The results from this performance assessment would be over-optimistic, and perhaps badly so.
To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
If the model was to be applied to adjacent agricultural regions, i.e., outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e. at the level of sub-regions within the learning sample, in order to realistically mimic the actual prediction task.
The CV resampling needed therefore depends as much on the prediction task itself as on the data structure or dependency at hand.

While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of appropriate spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
\pkg{mlr3spatiotempcv} is such a toolbox.

This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table \@ref(tab:sptcv-methods)).
The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on, for example, geometric blocking, clustering, and buffering.
In contrast, \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support for temporal prediction tasks.
The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
One limitation of these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework offers a seamless integration into model selection.
It is worth noting that a spatial cross-validation library named \pkg{spacv} has recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].

Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
This package is furthermore equipped with a variety of two- and three-dimensional visualization capabilities.
Our hope is that this implementation will facilitate reproducible geospatial modeling and code-sharing across a broad range of application domains.

The purpose of this paper is to give an overview of the methods implemented in the R package \pkg{mlr3spatiotempcv}.
After presenting the conceptual background in the following section, the overall structure of the \pkg{mlr3spatiotempcv} package is outlined.
Next, various spatial and spatiotemporal partitioning techniques are contrasted and compared, before their application is demonstrated in a machine-learning model assessment in the following section.
Finally, recommendations for the selection of suitable resampling techniques are given.

@@

In CV for predictive model assessment, we consider the following formal setting.
We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})^t\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
Our goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
\[
\mathit{perf(\hat{f}_\mathcal{L})} := E(l(Y,\hat{f}_\mathcal{L}(X))),
\]
where $l$ is a real-valued loss function, and the expected value is with respect to the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
This is referred to as the *actual* or *conditional* performance measure as it is conditional on $\mathcal{L}$ [@hand1997].
The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance measures for regression and classification (see, e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).

Since we only have a sample $\mathcal{T}$ of test data drawn from the population, we can only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
\[
\widehat{\mathit{perf}}_T(\hat{f}_\mathcal{L}) = \frac{1}{|\mathcal{T}|}\sum_{(Y,X)\in\mathcal{T}}l(Y,\hat{f}_\mathcal{L}(X)).
\]
This representation as a point estimator of $\mathit{perf(\hat{f}_\mathcal{L})}$ underlines the importance of using a random sample for model assessment to avoid estimation bias. Other estimators than the simple mean may be required when $\mathcal{T}$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson2012].
As always, judgment sampling may lead to uncontrollable bias.

Since re-using the learning sample $\mathcal{L}$ for testing, i.e. $\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, CV partitions the sample $\mathcal{L}$ into disjoint training and test sets.
Specifically, $\mathcal{L}$ is split into $k$ partitions,
\[
\mathcal{L} = \mathcal{L}_1 \cup \ldots \cup \mathcal{L}_k,\qquad \mathcal{L}_i\cap \mathcal{L}_j = \emptyset\quad \textrm{for all}\ i\neq j,
\]
and a model $\hat{f}_{(i)}$ is fitted on $\mathcal{L}_{(i)} := \mathcal{L}\setminus \mathcal{L}_i$, while $\mathcal{L}_i$ is withheld for testing.
This is repeated for $i=1,\ldots,k$ in order to effectively use the entire sample for testing, while keeping training and test sets disjoint at all times.
The $k$-fold CV estimator can therefore be written as
\[
\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f) := \frac{1}{k}\sum_{i=1}^k\widehat{\mathit{perf}}_{\mathcal{L}_i}(\hat{f}_{\mathcal{L}_{(i)}}),
\]
where $f$ is a machine-learning algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004; @cawley2010].
The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983].
Since $\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f)$ still depends on the particular partitioning chosen for $\mathcal{L}$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation) to reduce the influence of randomness when creating partitions [@vanwinckelen2012].

@@

Approaches based on *spatial blocking* (or *grouping*) require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
Let's refer to these spatial units or blocks as $\mathcal{Z}_i$, $1\le i\le n_z$.
These blocks are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010], which we refer to as *coordinate-based clustering*, or generating the desired number of rectangular blocks as an example of *geometric partitioning*.
The blocks may also be defined by a modeler based on an arbitrary partitioning of the study region based on an external data source, which we refer to as *custom resampling*.
This often used when the data is grouped. For example, when using to multi-level sampling designs or studying spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
We will broadly refer to such groups of observations as 'blocks' in a generic sense, regardless of the shape or origin of the groups.
Also, the data can be partitioned in feature space instead of geographic space, which has been referred to as "environmental blocking" [@roberts2017].

@@

A variant of CV is leave-one-out (LOO) CV, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
Although this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].

@@

With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip machine-learning frameworks such as \pkg{mlr3} with suitable algorithms.
In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extensible framework designed to accommodate numerous machine-learning tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.

With its integrative approach and its aim to provide long-term support, \pkg{mlr3} overcomes the challenges of combining multiple specialized packages with poorly standardized interfaces.
Issues that practitioners often face include varying argument lists of learners, different return values of `predict` methods, and support for only specific feature types.
These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the machine-learning pipelines.

Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., random CV); other specialized strategies (e.g., \texttt{TaskClassifST} or \texttt{TaskRegrST}) are defined in extension packages such as \pkg{mlr3spatiotempcv}.
In the machine-learning pipeline, these objects define the data splits used for model assessment and selection (hyperparameter tuning) by machine-learning algorithms.
Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@brenning2012; @blockCV; @cast], in order to expose them to \pkg{mlr3} functionality.
To reduce dependencies, some methods were re-implemented instead of importing them from upstream packages.

Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behavior from the underlying upstream R package.

Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods that visualize the groups created.
`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in three dimensions, i.e., geographic space plus time.

While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the mlr3 ecosystem to provide dedicated spatiotemporal learner and prediction methods.

@@

```{r rsmp-schema, echo=FALSE, fig.cap="Conceptual overview of various spatial partitioning schemas. Starting from unpartioned spatial observations (top left) either a spatial block partitioning or a 'spatial leave-one-out resampling' is applied in the first step. A spatial block partitioning can further be turned into a 'leave-one-block-out resampling' or a 'k-fold CV resampling at the block level'. The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.", out.width="100%",fig.pos="ht"}
knitr::include_graphics("pdf/rsmp-schema.pdf")
```

\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods) with their \pkg{mlr3} class name.
Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
The following sections give an overview of most implemented partitioning strategies and their visualization options.
The available methods are further discussed in section \@ref(sec:disc).

>>>>

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich Schiller University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    # must use German name, see https://www.lmu.de/de/die-lmu/struktur/zentrale-universitaetsverwaltung/kommunikation-und-presse/lmu-brand-guide/content-style-guide/index.html
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena \AND'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-Universität München'
    address: |
      | Department of Statistics
      | Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for machine learning in R with \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for machine learning in R with mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatial and spatiotemporal machine-learning models require a suitable framework for their model assessment, model selection, and hyperparameter tuning, in order to avoid error estimation bias and over-fitting.
  This contribution reviews the state-of-the-art in spatial and spatiotemporal CV, and introduces the R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) as an extension package of the machine-learning framework [mlr3](https://mlr3.mlr-org.com).
  Currently various R packages implementing different spatiotemporal partitioning strategies exist: blockCV, CAST, skmeans and sperrorest.
  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and common interface.
  This is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine-learning framework, which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
  One advantage is the use of a consistent nomenclature in an overarching machine-learning toolkit instead of a varying package-specific syntax, making it easier for users to choose from a variety of spatiotemporal resampling methods.
  This package avoids giving recommendations which method to use in practice as this decision depends on the predictive task at hand, the autocorrelation within the data, and the spatial structure of the sampling design or geographic objects being studied.
keywords:
  formatted: [cross-validation, predictive performance, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
  plain:     [cross-validation, predictive performance, autocorrelation, spatial, temporal, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{adjustbox}
  \usepackage{multirow}
  \usepackage{tabularx}
  \usepackage{placeins}
  \usepackage{scrextend}
  \usepackage{tablefootnote}
  \newcommand{\specialcell}[2][l]{%
    \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
tables: true
bibliography: paper.bib
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
---

@@

Although alternatives such as the bootstrap exist since some decades [@efron1983; @hand1997], cross-validation (CV) is a particularly well-established, easy-to-implement algorithm for *model assessment* of supervised machine-learning models [@efron1983 and next section] and *model selection* [@arlot2010].
In its basic form, CV is based on resampling the data without paying attention to any possible dependence structure, which may arise from, e.g., grouped or structured data, or underlying environmental processes inducing some sort of spatial coherence at the landscape scale.
In treating dependent observations as independent, or ignoring autocorrelation, CV test samples may in fact be heavily correlated with, or even pseudo-replicates of, the data used for training the model, which introduces a potentially severe bias in assessing the transferability of flexible machine-learning (ML) models.

This CV bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
It is most easily understood from a predictive modeling perspective by focusing on the question where (and when) the model should be used for prediction.
In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain [*intra-domain* prediction, @brenning2005] --- not *within* the same field, which obviously presents only a single crop type that is already known from the training sample.
In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and to test it on the remaining data, using other grid cells from the same fields, as if one wanted to predict within a field.
The results from this performance assessment would be over-optimistic, and perhaps badly so.
To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
If the model was to be applied to adjacent agricultural regions, i.e., outside the learning sample's spatial domain [*extra-domain* prediction, @brenning2005], it would even seem necessary to resample at a higher level of spatial aggregation, i.e. at the level of sub-regions within the learning sample, in order to realistically mimic the actual prediction task.
The CV resampling needed therefore depends as much on the prediction task itself as on the data structure or dependency at hand.

While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, the example from above may suffice to motivate the use of appropriate spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
\pkg{mlr3spatiotempcv} is such a toolbox.

This toolbox, implemented as an open-source R package, builds upon and generalizes several existing toolboxes that have been developed in recent years for more specific settings (Table \@ref(tab:sptcv-methods)).
The earliest and most comprehensive of these implementations is the \pkg{sperrorest} R package [@brenning2012], which provides an extensible framework and includes predefined resampling strategies based on geometric blocking, clustering, and buffering.
In contrast, packages \pkg{blockCV} and \pkg{ENMeval} were developed for block and buffer resampling with a focus on species distribution modeling [@blockCV; @rest2014; @muscarella2014].
However, neither of these have been integrated into established machine-learning frameworks such as \pkg{mlr}/\pkg{mlr3} or \pkg{caret}/\pkg{tidymodels}, and all of them lack support for temporal prediction tasks.
The \pkg{CAST} package, in contrast, focuses on spatiotemporal prediction tasks and makes use of some functions of the \pkg{caret} framework [@cast; @meyer2018].
One limitation of all these packages is the sole focus on model assessment, while the proposed implementation within the \pkg{mlr3} framework also offers seamless integration into model selection and provides parallel execution and enhanced logging abilities.
It is worth noting that a spatial cross-validation library named \pkg{spacv} has recently been developed for Python3, which can be used with the \pkg{scikit-learn} machine-learning framework [@pedregosa2011].

Thus, \pkg{mlr3spatiotempcv} implements for the first time a comprehensive state-of-the-art compilation of spatial and spatiotemporal partitioning schemes that is well-integrated into a comprehensive machine-learning framework in R, the \pkg{mlr3} ecosystem [@mlr3].
This package is furthermore equipped with a variety of two- and three-dimensional visualization capabilities.
The hope is that this implementation will simplify and facilitate reproducible geospatial modeling and code-sharing across a broad range of application domains.

The purpose of this article is to give an overview of the methods implemented in the R package \pkg{mlr3spatiotempcv}.
After presenting the conceptual background in the following section, the overall structure of the \pkg{mlr3spatiotempcv} package is outlined.
Next, various spatial and spatiotemporal partitioning techniques are contrasted and compared, before their application is demonstrated in a machine-learning model assessment in the following section.
Finally, recommendations for the selection of suitable resampling techniques are given.

@@

In CV for predictive model assessment, the following formal setting is considered.
The interest is in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})^t\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$.
The goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
\[
\mathit{perf(\hat{f}_\mathcal{L})} := E(l(Y,\hat{f}_\mathcal{L}(X))),
\]
where $l$ is a real-valued loss function, and the expected value is with respect to the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
This is referred to as the *actual* or *conditional* performance measure as it is conditional on $\mathcal{L}$ [@hand1997].
The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance measures for regression and classification (see, e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).

Since there is only a sample $\mathcal{T}$ of test data drawn from the population, one can only *estimate* the conditional performance of $\hat{f}_\mathcal{L}$:
\[
\widehat{\mathit{perf}}_T(\hat{f}_\mathcal{L}) = \frac{1}{|\mathcal{T}|}\sum_{(Y,X)\in\mathcal{T}}l(Y,\hat{f}_\mathcal{L}(X)).
\]
This representation as a point estimator of $\mathit{perf(\hat{f}_\mathcal{L})}$ underlines the importance of using a random sample for model assessment to avoid estimation bias. Other estimators than the simple mean may be required when $\mathcal{T}$ is not a simple random sample, for instance a stratified random sample [e.g., @thompson2012].
As always, judgment sampling may lead to uncontrollable bias.

Since re-using the learning sample $\mathcal{L}$ for testing, i.e. $\mathcal{T}:=\mathcal{L}$, would yield the over-optimistic *resubstitution* or *apparent* performance, CV partitions the sample $\mathcal{L}$ into disjoint training and test sets.
Specifically, $\mathcal{L}$ is split into $k$ partitions,
\[
\mathcal{L} = \mathcal{L}_1 \cup \ldots \cup \mathcal{L}_k,\qquad \mathcal{L}_i\cap \mathcal{L}_j = \emptyset\quad \textrm{for all}\ i\neq j,
\]
and a model $\hat{f}_{(i)}$ is fitted on $\mathcal{L}_{(i)} := \mathcal{L}\setminus \mathcal{L}_i$, while $\mathcal{L}_i$ is withheld for testing.
This is repeated for $i=1,\ldots,k$ in order to effectively use the entire sample for testing, while keeping training and test sets disjoint at all times.
The $k$-fold CV estimator can therefore be written as
\[
\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f) := \frac{1}{k}\sum_{i=1}^k\widehat{\mathit{perf}}_{\mathcal{L}_i}(\hat{f}_{\mathcal{L}_{(i)}}),
\]
where $f$ is a ML algorithm, i.e. a mapping that trains a model $\hat{f}_\mathcal{S}$ using any suitable training sample $\mathcal{S}$.
The use of $k=5$ or $k=10$ folds is most commonly seen in practice, and these preferences are also supported by theory [@bengio2004; @cawley2010].
The $k$-fold CV estimator of model performance is a nearly unbiased estimator of the conditional performance measure when the observations were drawn independently [@efron1983].
Since $\widehat{\mathit{perf}}_{\mathcal{L}, CV}(f)$ still depends on the particular partitioning chosen for $\mathcal{L}$, it is sometimes recommended to repeat the estimation using different random partitionings ($r$-repeated $k$-fold cross-validation) to reduce the influence of randomness when creating partitions [@vanwinckelen2012].

@@

Approaches based on *spatial blocking* (or somtimes calld *grouping*) require either the construction of spatial zones, or the use of pre-existing spatial structures in the data.
Let's refer to these spatial units or blocks as $\mathcal{Z}_i$, $1\le i\le n_z$.
These blocks are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010], which we refer to as *coordinate-based clustering*; or generating the desired number of rectangular blocks as an example of *geometric partitioning*.
The blocks may also be defined by a modeler based on an arbitrary partitioning of the study region based on an external data source, which we refer to as *custom resampling*.
This often used when the data is grouped. For example, when using to multi-level sampling designs or studying spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
We will broadly refer to such groups of observations as 'blocks' in a generic sense, regardless of the shape or origin of the groups.
Also, data can be partitioned in feature space instead of geographic space, which has been referred to as "environmental blocking" [@roberts2017].

@@

One variant of CV is leave-one-out (LOO) CV, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
Although this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].

@@

With the increased awareness of the importance of spatial and spatiotemporal resampling strategies and the growing popularity of R in environmental modeling and geocomputation, it is important to equip ML frameworks such as \pkg{mlr3} with suitable algorithms.
In this context, the \pkg{mlr3} ecosystem stands out as a unified, object-oriented and extensible framework designed to accommodate numerous ML tasks with a variety of learners, feature and model selection tools, and model assessment capabilities [@mlr3; @mlr3book].
All of these are supported by advanced visualization tools, which are particularly important in a spatial and spatiotemporal setting.

With its integrative approach and its aim to provide long-term support, \pkg{mlr3} overcomes the challenges of combining multiple specialized packages with poorly standardized interfaces.
Issues that practitioners often face include varying argument lists of learners, different return values of `predict()` methods, and support for only specific feature types.
These challenges result in substantial overhead and possible reproducibility issues, which are exacerbated by asynchronous development timelines of different components of the used ML pipelines.

Within the \pkg{mlr3} ecosystem, partitioning strategies are represented by their own objects of class \texttt{Resampling}, most of which are available within \pkg{mlr3} itself (e.g., random CV); other specialized strategies are defined in extension packages such as \pkg{mlr3spatiotempcv}.
In the ML pipeline, these objects define the data splits used for model assessment and selection (hyperparameter tuning) by ML algorithms.
Spatial and spatiotemporal partitioning techniques in \pkg{mlr3spatiotempcv} are currently mostly imported and interfaced from other packages, in particular \pkg{sperrorest}, \pkg{blockCV} and \pkg{CAST} [@brenning2012; @blockCV; @cast], in order to expose them to \pkg{mlr3} functionality.
To reduce dependencies, some methods were re-implemented instead of importing them from the respective upstream packages.

Resampling objects in \pkg{mlr3spatiotmpcv} inherit from class \texttt{mlr3::Resampling} and can be created from established object classes for geospatial data in R, including simple features [@pebesma2018], which facilitates their integration into domain-specific workflows in the geospatial sciences.
Support for projected (planar) and unprojected (geographic) coordinate reference systems (CRS) currently varies depending on the partitioning techniques used, since these inherit their behavior from the underlying upstream packages.

Partitioning objects in \pkg{mlr3spatiotempcv} are equipped with generic `plot()` and `autoplot()` methods to visualize the created partitions.
`autoplot()` is \pkg{ggplot2}-based and uses [ggplot2](https://ggplot2.tidyverse.org) [@ggplot2] in two-dimensional geographic space and [plotly](https://github.com/ropensci/plotly) [@plotly] in the three dimensions case, i.e., geographic space plus time.

While \pkg{mlr3spatiotempcv} solely focuses on spatiotemporal resampling methods and their visualization, other packages such as \pkg{mlr3spatial} or \pkg{mlr3temporal} are planned in the \pkg{mlr3} ecosystem to provide dedicated spatiotemporal learner and prediction methods.

@@

```{r rsmp-schema, echo=FALSE, fig.cap="Conceptual overview of various spatial partitioning schemas. Starting from unpartioned spatial observations (top left) either a 'spatial block partitioning' or a 'spatial leave-one-out resampling' is applied in the first step. A spatial block partitioning can further be turned into a 'leave-one-block-out resampling' or a 'k-fold CV resampling at the block level'. The use of a buffer is theoretically possible in any scenario but in practice only offered by specific method implementations.", out.width="100%",fig.pos="ht"}
knitr::include_graphics("pdf/rsmp-schema.pdf")
```

\pkg{mlr3spatiotempcv} currently implements the partitioning methods identified in Table \@ref(tab:sptcv-methods).
Several of the implemented algorithms are themselves versatile toolboxes with multiple options.
Comprehensive and up-to-date information can be found in the package's online documentation (<https://mlr3spatiotempcv.mlr-org.com>).
The following sections give an overview of most implemented partitioning strategies and their visualization options.
The available methods are further discussed in section \@ref(sec:disc).
